Can diffusion alone explain brain-wide distribution of CSF tracers within 24 hours?

### 2019-07-24 tex/rev1/paper.tex
updates

\subsection{Mathematical Model}
In \citet{sykova2008diffusion}, it was shown that the macroscopic diffusion in the brain can be considered a hindered diffusion with an apparent diffusion coefficient (ADC). The relation between the diffusion coefficients were defined as 
\begin{equation}
 \lambda =  \sqrt {D/D_{ADC}}
\label{tortuosity}
\end{equation}
with $\lambda$ denoted as the tortuosity.
In order to estimate the apparent diffusion coefficient involved in the contrast transportation shown in Figure~\ref{fig1} we assume that the process can be modeled by a diffusion equation. 
Then we constructed an optimization problem with the aim to minimize the difference between the observed and the modeled contrast distribution by optimizing the boundary conditions and the apparent diffusion coefficient. Thus enhanced transportation because of effects such as dissipation would result in an apparent diffusion coefficient larger than that predicted by DTI. The objective function was defined as 
\begin{equation}
\min_{D,g} \quad \sum\limits_{i=1}\sp{n} \int\limits_{\Omega} |u(t_i) - u_{obs}(t_i)|\sp{2} \mathrm{d}\Omega + \int\limits_{0}\sp{T} \int\limits_{\partial \Omega_1} \left( \frac{\alpha}{2} | g |\sp{2} + \frac{\beta}{2} \left| \frac{\partial g}{\partial t} \right|\sp{2} \right) \mathrm{d}\Omega \mathrm{d}t  
\label{EQ::objf}
\end{equation}
subject to   
\begin{equation}
\begin{aligned}
\frac{\partial u}{\partial t} &=  \nabla\sp{2} u && \text{in} \qquad \Omega \times \left\lbrace 0 , T \right)  \\
u&=g && \text{on} \qquad \partial\Omega_1  \times \left\lbrace 0 , T \right) 
\end{aligned}
\label{Eq::PDE}
\end{equation}
Here, $u$ is the contrast distribution, $D$ is the apparent diffusion 
coefficient, $g$ is the boundary condition, $\Omega$ is the domain, and $T$ is the final simulation time. We assume that the domain $\Omega$ consists of three sub domains, each with a different diffusion coefficient. We denote the CSF (subarachnoid and lateral ventricle) domain as $\Omega_1=\Omega_{CSF}$, the grey matter as $\Omega_{GM}$ and the white matter as $\Omega_{WM}$. 
The apparent
diffusion constant is assumed to be constant within the CSF, grey and 
white matter but each region may have different values.  
The Dirichlet boundary condition is only applied on the outer facing boundary of the CSF domain, $\partial \Omega_1$. Homogeneous Neumann conditions are applied on the remaining boundaries.
The $\alpha$ and $\beta$ parameters are non-negative regularization parameters 
and $u_{obs}$ are the observations at time-points $t_i$.

@@

\section{Numerical Experiments}
The process of estimating the diffusion coefficients in terms of the PDE constrained optimization problem \eqref{Eq::PDE} is complicated as appropriate parameters, i.e. $\alpha$, $\beta$ and $\gamma$,  have 
to be determined and robustness with respect to resolution and noise have to be assessed. Our approach was as follows.

@@

The implementation of the gradient regularization required that the boundary was decomposed to avoid the boundary values being continuous over different subdomains. We decomposed the boundary as seen in Figure~\ref{markedboundaryvalues}, with the red and blue boundary adjacent to the CSF. We defined the red boundary $\partial \Omega_r$ and the blue boundary $\partial \Omega_b$ as Dirichlet boundaries, while the green and yellow were Neumann boundaries. The regularization parameter $\gamma$ was added to the functional 
\begin{equation}
\min_{D,g} \quad \sum\limits_{i=1}\sp{n} \int\limits_{\Omega} |u(t_i) - u_{obs}(t_i)|\sp{2} \mathrm{d}\Omega + \int\limits_{0}\sp{T} \int\limits_{\partial \Omega_{rb}} \left( \frac{\alpha}{2} | g |\sp{2} + \frac{\beta}{2} \left| \frac{\partial g}{\partial t} \right|\sp{2}  + \frac{\gamma}{2}| \nabla g |\sp{2}  \right) \mathrm{d}\Omega \mathrm{d}t  
\end{equation}
We initially tested gradient regularization with the same parameter $\gamma$ on both boundaries, but the different distribution of tracers on the boundaries made it difficult to find an adequate value. It was observed that the concentration in the lateral ventricles were more uniform than that in the SAS, so we defined $\gamma$ as
\begin{equation}
\gamma = \begin{cases}
0.01\tilde{\gamma}  \in &\partial \Omega_{b} \\ 
\tilde{\gamma} \qquad \in  &\partial \Omega_{r}
\end{cases}  
\end{equation} 
with $\tilde{\gamma}$ as the referenced, i.e. mentioned in the text, regularization parameter.
The projection of values was implemented by taking the average of the adjacent CSF voxels to the boundary vertices on $\Omega_b$ and $\Omega_r$. We used the Gaussian smoothing found in the python-module scipy~\cite{jones2001scipy}, and applied the smoothing to the MRI-data. %The projection and smoothing approach changes the observations, while the gradient regularization adds additional term to the functional. Therefore a dual method can also be considered, 
%It can be seen from Fig XX that the boundary conditions is noise. This may be attributed to noise or small errors
%in the segmentation/registration process where voxels close to the boundaries between the brain and CSF are miss-aligned
%by the sub-voxel segementation procedure. In particular, the surfaces can cut a voxel at any plane, not only the planes
%through the voxel vertices. We chose three approaches to handle this problem. \kam{Lars: kan du utbrodere litt her? 
%Jeg synes vi skal beholde også Fig 10 fra tidligere.  }

@@

\hline
\end{tabular}
\caption{The table shows an overview of the parameters used in case of boundary constrained optimization with MRI data. }
\label{model-params-overview}
\end{table}

@@

\section{Insert correctly}
(The sampling of) The observations on the boundary between CSF and the brain parenchyma suffered from the poor resolution of the T1-map, which lead to high frequency noise on the boundary. Therefore we investigates the effect of a gradient regularization parameter, Gaussian smoothing and the projecting the average concentration onto the boundary has ways to reduce the noise.

The implementation of the gradient regularization required that the boundary was decomposed to avoid the boundary values being continuous over different subdomains. We decomposed the boundary as seen in Figure ~\ref{markedboundaryvalues}, with the red and blue boundary next to the CSF. We defined the red boundary $\partial \Omega_r$ and the blue boundary $\partial \Omega_b$ as Dirichlet boundaries, while the green and yellow were Neumann boundaries. The regularization parameter $\gamma$ was added to the functional 
\begin{equation}
\min_{D,g} \quad \sum\limits_{i=1}\sp{n} \int\limits_{\Omega} |u(t_i) - u_{obs}(t_i)|\sp{2} \mathrm{d}\Omega + \int\limits_{0}\sp{T} \int\limits_{\partial \Omega_{rb}} \left( \frac{\alpha}{2} | g |\sp{2} + \frac{\beta}{2} \left| \frac{\partial g}{\partial t} \right|\sp{2}  + \frac{\gamma}{2}| \nabla g |\sp{2}  \right) \mathrm{d}\Omega \mathrm{d}t  
\end{equation}
The boundaries $\Omega_r$ and $\Omega_b$ have very different shapes, and using the same regularization parameter can have different effects on the two boundaries. Based on the  observations we saw that that the concentration in the lateral ventricles were more uniform than that in the SAS. Therefore, we defined $\gamma$ as
\begin{equation}
\gamma = \begin{cases}
0.01\tilde{\gamma}  \in &\partial \Omega_{b} \\ 
\tilde{\gamma} \qquad \in  &\partial \Omega_{r}
\end{cases}  
\end{equation} 
with $\tilde{\gamma}$ as the referenced, i.e. mentioned in the text, regularization parameter.

The results for the ventricular boundary in Figure ~\ref{VENT} and the SAS boundary in Figure ~\ref{SAS}. The corresponding estimated diffusion coefficient are shown in Figure ??. 
We see that the gradient regularization has a great effect reducing the noise on the boundaries, but it also changed the computed diffusion coefficients.

The Gaussian smoothing was done with the python module scipy, and the standard deviation examined in the range of 0.3-1.5. We observed that the Gaussian smoothing with a standard deviation of 1.5 increased the computed diffusion coefficient by approximately 200\% compared to raw data.

>>>>

\subsection{Mathematical Model}
In \citet{sykova2008diffusion}, it was shown that the macroscopic diffusion in the brain can be considered a hindered diffusion with an apparent diffusion coefficient (ADC). The relation between the diffusion coefficients were defined as 
\begin{equation}
 \lambda =  \sqrt {D/D_{ADC}}
\label{tortuosity}
\end{equation}
with $\lambda$ denoted as the tortuosity.
In order to estimate the apparent diffusion coefficient involved in the contrast transportation shown in Figure~\ref{fig1} we assume that the process can be modeled by a diffusion equation. 
Then we constructed an optimization problem with the aim to minimize the difference between the observed and the modeled contrast distribution by optimizing the boundary conditions and the apparent diffusion coefficient. Thus enhanced transportation because of effects such as dissipation would result in an apparent diffusion coefficient larger than that predicted by DTI. The objective function was defined as 
\begin{equation}
\min_{D,g} \quad \sum\limits_{i=1}\sp{n} \int\limits_{\Omega} |u(t_i) - u_{obs}(t_i)|\sp{2} \mathrm{d}\Omega + \int\limits_{0}\sp{T} \int\limits_{\partial \Omega_1} \left( \frac{\alpha}{2} | g |\sp{2} + \frac{\beta}{2} \left| \frac{\partial g}{\partial t} \right|\sp{2} +  \frac{\gamma}{2}| \nabla g |\sp{2} \right) \mathrm{d}\Omega \mathrm{d}t  
\label{EQ::objf}
\end{equation}
subject to   
\begin{equation}
\begin{aligned}
\frac{\partial u}{\partial t} &=  \nabla\sp{2} u && \text{in} \qquad \Omega \times \left\lbrace 0 , T \right)  \\
u&=g && \text{on} \qquad \partial\Omega_1  \times \left\lbrace 0 , T \right) 
\end{aligned}
\label{Eq::PDE}
\end{equation}
Here, $u$ is the contrast distribution, $D$ is the apparent diffusion 
coefficient, $g$ is the boundary condition, $\Omega$ is the domain, and $T$ is the final simulation time. We assume that the domain $\Omega$ consists of three sub domains, each with a different diffusion coefficient. We denote the CSF (subarachnoid and lateral ventricle) domain as $\Omega_1=\Omega_{CSF}$, the grey matter as $\Omega_{GM}$ and the white matter as $\Omega_{WM}$. 
The apparent
diffusion constant is assumed to be constant within the CSF, grey and 
white matter but each region may have different values.  
The Dirichlet boundary condition is only applied on the outer facing boundary of the CSF domain, $\partial \Omega_1$. Homogeneous Neumann conditions are applied on the remaining boundaries.
The $\alpha$, $\beta$ and $\gamma$ parameters are non-negative regularization parameters 
and $u_{obs}$ are the observations at time-points $t_i$.

@@

%\section{Numerical Experiments}
The process of estimating the diffusion coefficients in terms of the PDE constrained optimization problem \eqref{Eq::PDE} is complicated as appropriate parameters, i.e. $\alpha$, $\beta$ and $\gamma$,  have 
to be determined and robustness with respect to resolution and noise have to be assessed. Our approach was as follows.

@@

The implementation of the gradient regularization required that the boundary was decomposed to avoid the boundary values being continuous over different subdomains. We decomposed the boundary as seen in Figure~\ref{markedboundaryvalues}, with the red and blue boundary adjacent to the CSF. We defined the red boundary $\partial \Omega_r$ and the blue boundary $\partial \Omega_b$ as Dirichlet boundaries, while the green and yellow were Neumann boundaries. The regularization parameter $\gamma$ was subsequentally to non-zero in Eq.~\ref{\label{EQ::objf}}. We initially tested gradient regularization with the same parameter $\gamma$ on both boundaries, but the different distribution of tracers on the boundaries made it difficult to find an adequate value. It was observed that the concentration in the lateral ventricles were more uniform than that in the SAS, so we defined $\gamma$ as
\begin{equation}
\gamma = \begin{cases}
0.01\tilde{\gamma}  \in &\partial \Omega_{b} \\ 
\tilde{\gamma} \qquad \in  &\partial \Omega_{r}
\end{cases}  
\end{equation} 
with $\tilde{\gamma}$ as the referenced, i.e. mentioned in the text, regularization parameter.

The projection method worked by selecting each boundary vertex on $\Omega_b$ and $\Omega_r$, then finding the corresponding voxel using an affine transformation. Then for each voxels, we  computed the average of all CSF segmented voxels in a two voxel vicinity. The averages was then set to be the observational value for corresponding boundary vertex in the computations.

We used the Gaussian smoothing found in the python-module scipy~\cite{jones2001scipy}, and applied the smoothing to the voxel based MRI-data. The smoothing function required that standard deviation of a Gaussian distribution was specified, which was tested in the range 0.3-1.5.

 %The projection and smoothing approach changes the observations, while the gradient regularization adds additional term to the functional. Therefore a dual method can also be considered, 
%It can be seen from Fig XX that the boundary conditions is noise. This may be attributed to noise or small errors
%in the segmentation/registration process where voxels close to the boundaries between the brain and CSF are miss-aligned
%by the sub-voxel segementation procedure. In particular, the surfaces can cut a voxel at any plane, not only the planes
%through the voxel vertices. We chose three approaches to handle this problem. \kam{Lars: kan du utbrodere litt her? 
%Jeg synes vi skal beholde også Fig 10 fra tidligere.  }

@@

Gradient Regularization   & $\beta$	   &	 $\lbrace 0.0, 100.0\rbrace$	\\ 
\end{tabular}
\caption{The table shows an overview of the parameters used in case of boundary constrained optimization with MRI data. }
\label{model-params-overview}
\end{table}

@@

%\section{Insert correctly}
%(The sampling of) The observations on the boundary between CSF and the brain parenchyma suffered from the poor resolution of the T1-map, which lead to high frequency noise on the boundary. Therefore we investigates the effect of a gradient regularization parameter, Gaussian smoothing and the projecting the average concentration onto the boundary has ways to reduce the noise. 
%
%The implementation of the gradient regularization required that the boundary was decomposed to avoid the boundary values being continuous over different subdomains. We decomposed the boundary as seen in Figure ~\ref{markedboundaryvalues}, with the red and blue boundary next to the CSF. We defined the red boundary $\partial \Omega_r$ and the blue boundary $\partial \Omega_b$ as Dirichlet boundaries, while the green and yellow were Neumann boundaries. The regularization parameter $\gamma$ was added to the functional 
%\begin{equation}
%\min_{D,g} \quad \sum\limits_{i=1}\sp{n} \int\limits_{\Omega} |u(t_i) - u_{obs}(t_i)|\sp{2} \mathrm{d}\Omega + \int\limits_{0}\sp{T} \int\limits_{\partial \Omega_{rb}} \left( \frac{\alpha}{2} | g |\sp{2} + \frac{\beta}{2} \left| \frac{\partial g}{\partial t} \right|\sp{2}  + \frac{\gamma}{2}| \nabla g |\sp{2}  \right) \mathrm{d}\Omega \mathrm{d}t  
%\end{equation}
%The boundaries $\Omega_r$ and $\Omega_b$ have very different shapes, and using the same regularization parameter can have different effects on the two boundaries. Based on the  observations we saw that that the concentration in the lateral ventricles were more uniform than that in the SAS. Therefore, we defined $\gamma$ as
%\begin{equation}
%\gamma = \begin{cases}
%0.01\tilde{\gamma}  \in &\partial \Omega_{b} \\ 
%\tilde{\gamma} \qquad \in  &\partial \Omega_{r}
%\end{cases}  
%\end{equation} 
%with $\tilde{\gamma}$ as the referenced, i.e. mentioned in the text, regularization parameter.
%
%The results for the ventricular boundary in Figure ~\ref{VENT} and the SAS boundary in Figure ~\ref{SAS}. The corresponding estimated diffusion coefficient are shown in Figure ??. 
%We see that the gradient regularization has a great effect reducing the noise on the boundaries, but it also changed the computed diffusion coefficients.
%
%
%The Gaussian smoothing was done with the python module scipy, and the standard deviation examined in the range of 0.3-1.5. We observed that the Gaussian smoothing with a standard deviation of 1.5 increased the computed diffusion coefficient by approximately 200\% compared to raw data.   
%
%
